---
title: "Written Report"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dev.args = list(bg = "transparent"),
  comment = NA
  )
```

```{r}
library(tidyverse)
library(knitr)
library(lubridate)
library(ggcorrplot)
library(plotly)
library(scales)
library(randomForest)
library(caTools)

steam_theme <- theme_classic() + 
  theme(
    panel.background = element_rect(fill = "transparent", color = NA), 
    plot.background = element_rect(fill = "transparent", color = NA),
    legend.background = element_rect(fill = "transparent", color = NA),
    legend.box.background = element_rect(fill = "transparent", color = NA),
    text = element_text(color = "#66c0f4")
    )

theme_set(steam_theme)
```

# Motivation

Video game industry is a highly competitive market where thousands of new products are released annually. For developers, understanding the specific factors that drive player engagement is important when allocating their resources. For players facing an overwhelming number of choices, identifying games that align with their interests is a challenge. This project is motivated by both the need to bridge this gap and our enthusiasm towards gaming: by analyzing real-world data from [Steam](https://en.wikipedia.org/wiki/Steam_(service)), we aim to build a predictive model that can estimate a game's popularity based on current metadata.

# Related Work

Our work is inspired by the recent kick-off of [TGA](https://thegameawards.com/nominees/game-of-the-year): An once-a-year celebration of game developers and players, simply consider it as the Oscar of gaming industry. Observing shifts in genre trends and consumer interest prompted us to explore quantitative patterns underlying game popularity.

# Initial Questions

Our major questions is: **What are the primary drivers of game popularity on Steam?**

As we explored the dataset, this question evolved into some hypotheses:

- "Does a higher price point correlate with higher quality and thus more players?" 
- "Are buyers necessarily players?"

- "Are games with more positive reviews being played more?"

- "Which types of games are being played the most? Action or RPG?"

- "Does a DLCs attract more players into playing?"

- "Does price range—not just price—relate to engagement?"

- "Do developers with more releases also create more popular games 

# Data

## Data Source

The dataset used for this project is made possible by [FronkonGames](https://fronkongames.github.io/) and is sourced at [huggingface.co](https://huggingface.co/datasets/FronkonGames/steam-games-dataset).

## Scraping

The author provides an official python script to scrape this dataset, it can be viewed [here](data/scrapping.py). Additionally, we did some sampling 
from local. The chunk below converted a dataframe of 110000 rows to 10629 rows to pass Github's file uploading size. It is not meant to be ran.
```{r}
# df = read_csv("steam_games.csv") |>
#   filter(!is.na(average_playtime),
#          !is.na(peak_ccu),
#          !is.na(median_playtime))

# write_csv(df, "steam_games.csv")
```

## Cleaning

During the cleaning process, our pipeline aims to:

- Make sure variable names are consistent across columns

- Select only the unique identifier and necessary variables for our popularity analysis

- Make sure continuous variables are numeric

- Make sure `release_date` is in date format for later manipulations

- Convert the `estimated_owners` from a range to a numeric value by taking the midpoint

- Create a `primary_genre` variable. This will be the first string in list `genres`

- Make sure no zeros or NAs are present in `average_playtime`, `median_playtime` and `peak_ccu` as they are meaningless in these columns

```{r}
steam_df = read_csv("data/steam_games.csv") |>
  janitor::clean_names() |>
  select(
    app_id, name, release_date, price,
    estimated_owners, peak_ccu, average_playtime, median_playtime,
    dlc_count, developers, genres, positive, negative
  ) |>
  mutate(
    price = as.numeric(price),
    dlc_count = as.numeric(dlc_count),
    positive = as.numeric(positive),
    negative = as.numeric(negative),
    release_date = mdy(release_date),
    primary_genre = as.factor(sub(",.*", "", genres)), 
  ) |>
  mutate(estimated_owners = str_extract(estimated_owners, "\\d+"),
         estimated_owners = as.numeric(estimated_owners)) |>
  filter(
    average_playtime > 0,
    median_playtime > 0,
    peak_ccu > 0
  )
```

## Key Variables

**app_id**: Unique identifier of each game

**name**: Game name on Steam

**release_date**: Official releasing time in yyyy-mm-dd

**price**: Price in USD

**estimated_owners**: Rough estimate of owners

**peak_ccu**: Peak number of online concurrent users

**average_playtime, median_playtime**: Playtime since March 2009, in minutes

**dlc_count**: Number of DLCs (expansion or extra chapters sold separately from the game)

**developers**: Creator of the game

**genres**: Theme of the game

**positive, negative**: Number of positive / negative reviews by user on Steam

# Exploratory Data Analysis

In our exploratory data analysis, we want to understand what factors drive game popularity on Steam. We begin by exploring the distribution of the key outcome variable: `peak_ccu` (peak concurrent users) along with their relationships to potential predictors such as playtime, price, release year, estimated owners, and genre/category information. In particular, we examine not only cross-sectional patterns (e.g., owners, price, reviews), but also temporal trends and engagement-related quantities such as release year and average playtime.

Specifically, our EDA aims to:

- Identify general patterns in popularity  
- Assess data quality and transformations needed for modeling  
- Detect early signals of which features might influence engagement over time and across different levels of player commitment


## Distribution of Peak Concurrent Users

```{r}
pccu_distribution = steam_df |>
  ggplot(aes(x = peak_ccu)) +
  geom_histogram(bins = 50, fill = "#66c0f4") +
  scale_x_continuous(trans = "log10") +
  labs(
    title = "Distribution of Peak Concurrent Users (Log Scale)",
    x = "Peak CCU"
  )

ggplotly(pccu_distribution)
```

Note that peak_ccu is heavily tailed. A tiny handful of games (like Counter-Strike 2, Dota 2, PUBG) have millions of CCU and owners, while the vast majority of games have very few owners and almost zero peak CCU. Applying log scales here allow us to compare games based on multiplicative differences rather than additive ones. For developers, the jump from 1,000 owners to 10,000 owners is just as significant as the jump from 100,000 to 1,000,000.

## Estimated Owners vs Popularity 

Owners represent the overall reach of a game while peak_ccu reflects active use. Games with high owner counts but low CCU may indicate weak long-term engagement. The plot below visualizes the correlation between estimated owners and peak CCU.

```{r}
eo_boxplot = steam_df |>
  ggplot(aes(x = factor(estimated_owners), y = peak_ccu)) +
  geom_boxplot(fill = "#66c0f4", alpha = 0.6, color = "#66c0f4") +
  scale_y_continuous(trans = "log10") +
  labs(
    title = "Distribution of Peak CCU by Owner Tier",
    x = "Estimated Owners Tier",
    y = "Peak CCU (Log Scale)"
  )

ggplotly(eo_boxplot)
```

In the lower tiers (left side), the whiskers are long, and outliers are everywhere. Performance is unpredictable. In the highest tiers ($1e+07$), the boxes are tighter. Namely, If you have 10 million owners, your CCU is predictably massive. There are very few "dead" games with 10 million owners.

# Free vs Paid Games

Free-to-play titles often follow different business models than paid games.Free-to-play games often mean a larger player base and higher expectations for popularity fluctuations, do free-to-play games have a higher CCU than paid games? Here we compare the distribution of popularity between free and paid games.

```{r,message=FALSE, warning=FALSE}
steam_free_paid = steam_df |>
  mutate(
    free_flag = if_else(price == 0, "Free", "Paid")
  )
free_paid_plot = steam_free_paid |>
  ggplot(aes(x = free_flag, y = peak_ccu)) +
  geom_boxplot(fill = "#66c0f4", alpha = 0.6, color = "#66c0f4") +
  scale_y_continuous(trans = "log10") +
  labs(
    title = "Free vs Paid Games Peak CCU ",
    x = "",
    y = "Peak CCU (log scale)"
  )

ggplotly(free_paid_plot)
```

The median peak CCU is similar for both free-to-play and paid games, indicating that there is no fundamental difference in popularity between typical free-to-play games and typical paid games.However, Free-to-play games have higher dispersion and more hit games. The longer upper bounds and more extreme outliers show that most of the most popular games on Steam (such as PUBG, CS2, and Apex Legends) are free-to-play. Paid games are more concentrated. Paid games can also be very popular, but their overall distribution is more compact, with less fluctuation in popularity

# Price Range and Game Popularity

Based on the above analysis, we further divide the price of paid games into different ranges to observe the median CCU at different price levels.

```{r ,message=FALSE, warning=FALSE}
steam_price_level <- steam_df |>
  mutate(
    price_level = case_when(
      price <= 4.99 ~ "$0.99–4.99",
      price <= 9.99 ~ "$5–9.99",
      price <= 19.99 ~ "$10–19.99",
      price <= 39.99 ~ "$20–39.99",
      price <= 59.99 ~ "$40–59.99",
      TRUE ~ "$60+"
    )
  )

price_median_plot <- steam_price_level |>
  group_by(price_level) |>
  summarize(
    median_ccu = median(peak_ccu),
    .groups = "drop"
  ) |>
  ggplot(aes(x = reorder(price_level, median_ccu), y = median_ccu)) +
  geom_col(fill = "#66c0f4", alpha = 0.8) +
  coord_flip() +
  scale_y_continuous(trans = "log10") +
  labs(
    title = "Median Peak CCU by Price Range",
    x = "Price Range",
    y = "Median Peak CCU (log scale)"
  )

ggplotly(price_median_plot)
```

We can see that the more expensive the game, the higher the median Peak CCU.This phenomenon isn't because players prefer expensive games, but rather because higher prices usually represent large-scale productions . These games are more competitive in terms of marketing and production quality, thus attracting more players. Conversely, games priced below $10 are typically indie games with a smaller target audience and limited market exposure, resulting in a naturally lower median popularity.

## User Satisfaction vs Popularity

Reviews reflect user satisfaction. If higher satisfaction correlates with higher playtime, review sentiment becomes a strong candidate predictor. Here, we define satisfaction with Review Ratio:
\[
RR = \frac{Positive}{Positive + Negative} 
\]

```{r}
rr_trend = steam_df |>
  mutate(
    total_reviews = positive + negative,
    review_ratio = positive / total_reviews
  ) |>
  filter(total_reviews > 50) |>
  ggplot(aes(x = review_ratio, y = peak_ccu)) +
  geom_point(alpha = 0.2, size = 0.8, color = "#66c0f4") +
  geom_smooth(method = "gam", color = "#e6e9f0", se = FALSE) +
  scale_y_log10(
    labels = label_number(accuracy = 1, big.mark = ","),
    breaks = trans_breaks("log10", function(x) 10^x)
  ) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Review Ratio vs Peak CCU",
    x = "Positive Review Ratio",
    y = "Peak CCU"
  )

ggplotly(rr_trend)
```

The white trend line stays relatively flat for a long time and then curves upward sharply once the Review Ratio exceeds roughly 75%.
This implies that moving from a "Bad" game (25%) to a "Mediocre" game (50%) yields very little gain in popularity. However, moving from "Good" (75%) to "Excellent" (95%+) yields an significant increase in peak players.


## Popularity by Genre

Genres encapsulate game design choices, and identifying which genres drive deeper engagement is important for game developers. Below we list the top 15 most popular genres on Steam.

```{r}
genres_trend_plot = steam_df |>
  separate_rows(genres, sep = ",\\s*") |>
  filter(!genres %in% c("Free to Play", "Early Access")) |>
  group_by(genres) |>
  summarize(
    total_ccu = sum(peak_ccu, na.rm = TRUE),
    n_games = n()
  ) |>
  arrange(desc(total_ccu)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(genres, total_ccu), 
             y = total_ccu,
             text = paste("Genre:", genres, 
                          "<br>Total CCU:", scales::comma(total_ccu),
                          "<br>Number of Games:", n_games))) +
  geom_col(fill = "#66c0f4") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Top Genres by Total Peak CCU",
    x = "Genre",
    y = "Total Peak CCU"
  )

ggplotly(genres_trend_plot, tooltip = "text")
```

The plot above gives very intuitive results, and the 5 most popular genres are: Action, Adventure, RPG(Role Play), Indie, and Simulation. Clicking on each bar will show the number of games in that genre. For developers, a bigger market (more CCU) is almost certainly filled with more competitors. RPG is a very special genre with 6,329,228 peak CCU but only 2590 games, as the cost of making a RPG is usually expensive.

# Developer Output and Productivity

Steam developers vary widely in scale, some studios release dozens of games, while others release only one. Here we examine:
1. Which developers publish the most games  
2. Which developers achieve the highest typical popularity 

```{r, message=FALSE, warning=FALSE}

dev_df <- steam_df |>
  separate_rows(developers, sep = ",\\s*")

dev_count_plot <- dev_df |>
  count(developers) |>
  arrange(desc(n)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(developers, n), y = n)) +
  geom_col(fill = "#66c0f4") +
  coord_flip() +
  labs(
    title = "Top 15 Developers by Number of Games Published",
    x = "Developer",
    y = "Number of Games"
  )

ggplotly(dev_count_plot)

dev_quality_plot <- dev_df |>
  group_by(developers) |>
  summarize(
    n_games = n(),
    median_ccu = median(peak_ccu)
  ) |>
  filter(n_games >= 5) |>  
  arrange(desc(median_ccu)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = reorder(developers, median_ccu), y = median_ccu)) +
  geom_col(fill = "#66c0f4") +
  coord_flip() +
  scale_y_continuous(trans = "log10") +
  labs(
    title = "Top Developers by Median Peak CCU ",
    x = "Developer",
    y = "Median Peak CCU (log scale)"
  )

ggplotly(dev_quality_plot)
```

Our developer analysis reveals a clear divide between quantity-driven and quality-driven production on Steam. Developers that publish the largest number of titles—many of which appear under generic labels such as “Inc.” or “Ltd.”—do not correspond to the creators of the most popular games. In contrast, studios with the highest median peak concurrency, such as Nixxes Software, Insomniac Games, and Respawn Entertainment, are well-established AAA teams whose success stems from high production value, strong IP franchises, and consistent player engagement. This contrast highlights that the number of games released is not a predictor of player popularity; instead, sustained quality and brand strength are what drive meaningful player traffic on the platform.

## Correlation

Lastly, we use a correlation graph to check for potential predictors for our model.

```{r}
correlation_matrix = steam_df |>
  mutate(
    release_date = as.numeric(release_date),
    total_reviews = positive + negative,
    review_ratio = ifelse(total_reviews > 0, positive / total_reviews, 0),
    log_peak_ccu = log1p(peak_ccu),
    log_estimated_owners = log1p(estimated_owners),
    log_price = log1p(price),
    log_average_playtime = log1p(average_playtime)
  ) |>
  select(log_price, log_estimated_owners, release_date,
         log_peak_ccu, log_average_playtime, median_playtime,
         dlc_count, review_ratio) |>
  cor(use = "complete.obs") |>
  ggcorrplot(
    lab = FALSE, 
    hc.order = TRUE, 
    ggtheme = steam_theme,
    outline.color = "white",
    colors = c("#6D9EC1", "white", "#E46726"))

ggplotly(correlation_matrix)
```

From the correlation matrix and other plots, we find `peak_ccu` and most variables are weak linearly. Therefore, for our additional analysis, instead of performing a Linear Regression, we use a Random Forest regressor to unveil relationships and interactions.

# Random Forest

## Modeling Strategy

This additional analysis aims to predict the maximum concurrent player count (`peak_ccu`) for Steam games based on current metadata. By modeling the relationship between game attributes—such as price, genre, and release timing—and player traffic, we can identify the key drivers of a game's commercial performance.

We selected a Random Forest as our modeling algorithm. This decision was made by the some characteristics of the Steam dataset observed during [Exploratory Data Analysis](eda.html):

- **Weak Linearity**: Our correlation matrix revealed that relationships between variables are barely linear. Random Forest captures complex, non-linear interactions without requiring manual equation fitting.

- **Robustness to Outliers**: Steam data has a skewed distribution, where a few massive hits coexist with thousands of small games. Random Forest averages the results of multiple decision trees, preventing these extreme outliers from messing up the entire model.

Before fitting the Random Forest, we added a baseline:
Baseline Model : Linear Regression
This model assumes a linear relationship between features and log(Peak CCU).
It serves as a benchmark to quantify Random Forest’s non-linear improvement.

## Data Preparation

To address the extreme skewness of the data, we perform a log1p transformation ($y = \ln(x + 1)$) to `peak_ccu` and continuous features. This transformation, as we previously mentioned, allows the model to learn magnitudes.

```{r}
model_data = steam_df |>
  mutate(
    log_peak_ccu = log1p(peak_ccu),
    log_price = log1p(price),
    release_date = as.numeric(release_date),
    total_reviews = positive + negative,
    review_ratio = ifelse(total_reviews > 0, positive / total_reviews, 0),
    log_estimated_owners = log1p(estimated_owners),
    log_average_playtime = log1p(average_playtime)
  ) |>
  select(log_peak_ccu, log_price, log_estimated_owners, log_average_playtime, dlc_count, primary_genre, release_date, review_ratio) |>
  na.omit()
```

## Training Regressoor

We first set a seed to make sure our work is reproducible. The dataset was then split into training (80% of original data) and testing (20% of original data) sets. We then generated a basic model with only individual variables as a control group. Finally, we start training our Random Forest regressor with 500 decision trees and save this model as `rf_model`. The details of basic model and `rf_model` is given below.
```{r}
set.seed(8105)

split = sample.split(model_data$log_peak_ccu, SplitRatio = 0.8)
train_set = subset(model_data, split == TRUE)
test_set = subset(model_data, split == FALSE)

baseline_model <- lm(log_peak_ccu ~ ., data = train_set)

rf_model = randomForest(log_peak_ccu ~ ., 
                         data = train_set, 
                         ntree = 500, 
                         importance = TRUE)

print(rf_model)
```

## Evaluation


```{r}
valid_genres <- unique(train_set$primary_genre)
test_setnew <- test_set |>
  dplyr::filter(primary_genre %in% valid_genres)
log_pred_lm <- predict(baseline_model, newdata = test_setnew)
log_pred_rf <- predict(rf_model,     newdata = test_setnew)
real_actuals <- expm1(test_setnew$log_peak_ccu)
pred_lm      <- expm1(log_pred_lm)
pred_rf      <- expm1(log_pred_rf)

rmse_fun <- function(pred, obs) {
  sqrt(mean((pred - obs)^2))
}

lm_rmse <- rmse_fun(pred_lm, real_actuals)
rf_rmse <- rmse_fun(pred_rf, real_actuals)

lm_nrmse <- lm_rmse / mean(real_actuals)
rf_nrmse <- rf_rmse / mean(real_actuals)

results_table <- tibble::tibble(
  Model = c("Linear Regression", "Random Forest"),
  RMSE  = c(lm_rmse, rf_rmse),
  NRMSE = c(lm_nrmse, rf_nrmse)
)

knitr::kable(results_table, digits = 4)
```

We first generate the log-scale predictions. Then we reverse our log transformation with exponential equation to get the RMSE (Root Mean Squared Error) in actual player units. Additionally, we apply normalization to our RMSE to get NRMSE, where:
\[
NRMSE = \frac{RMSE}{Mean}
\]

NRMSE tells us the error percentage relative to the average game. The baseline model gives a RMSE of 142.43 and NRMSE of 2.99, and final model achieved a RMSE of 127.38 and a Normalized RMSE of 2.67. Considering the large variance of the dataset, our model holds a fairly stable predictive capability.

# Result & Discussion

## Findings
From the Variable Importance Plot of our model below, there are several findings:

```{r}
imp_dat <- as.data.frame(importance(rf_model))
imp_dat$var_name <- rownames(imp_dat)

ggplot(imp_dat, aes(x = IncNodePurity, y = reorder(var_name, IncNodePurity))) +
  geom_point(color = "#66c0f4", size = 3) +
  geom_segment(aes(x = 0, xend = IncNodePurity, y = var_name, yend = var_name),
               color = "#66c0f4") + 
  labs(title = "What drives Peak CCU?",
       x = "Importance (IncNodePurity)",
       y = "Variable Name")
```

- **Fans of Recency (`release_date`)**: The most significant predictor of peak_ccu was the release date. This indicates a strong "recency bias" in the Steam marketplace: people prefer new stuff. Newer games benefit from the "hype cycle" and marketing pushes, while older titles—regardless of quality—tend to see significant player churn. The model successfully learned the lifecycle decay of a typical video game.

- **Buyers are Players (`estimated_owners`)**: The number of estimated owners served as a strong secondary predictor. However, this result may come from the logical constraint that people can only play a game when they own it.

- **The Role of Content (`primary_genre`)**: Surprisingly, primary_genre showed lower predictive power compared to other factors. This result suggests that while genre defines the type of player, it does not strictly dictate the volume of players. This may also reflect a limitation in our feature engineering: by isolating only the primary genre, the model may miss the nuances of multi-genre titles.

## Conclusion

Our model validates that timing and reach (release window and ownership) are statistically more significant predictors of peak traffic than price or broad genre classifications. The fact that these two predictors shows in top is very surprising, as none of the team members expected or proposed such outcome. Future iterations of this study could improve accuracy by incorporating granular tag analysis (e.g., "Co-op", "PvP") to better capture the retention mechanics of specific game loops.
